Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-03-02 manuscript/manuscript.tex
small tweaks to compete section

We can observe this phenomenon of shifting amino-acid preferences degrading the accuracy of the ExpCM by fitting trees containing increasingly diverged sequences.
Specifically, for each HA for which we have deep mutational scanning data, we created three HA phylogenetic trees: a ``low'' divergence tree that contains only sequences with $\ge ?$ amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with $\ge ?$ amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all the HAs (which have as little as 38\% identity to the HA in the experiment).
\ref{fig:compete} shows these sets of HA sequences.
For each of the subtrees in \ref{fig:compete}, we examined the congruence between site-specific natural selection over the tree and the amino-acid preferences measured in the deep mutational scanning experiment. 
\jdbcomment{Some general description of $\beta$ here. I think we'd noted above possibly describing it then? If so, just put a short recap here. If you decide not to describe it above in any detail, have something longer here.}
The fit $\beta$ parameter rescales the raw preferences from the deep mutational scan and relates the selection in nature to selection in the lab. 
We interpret $\beta >1$ as selection in nature favoring the same amino acids as selection in the experiments, or that the amino-acid preferences are a good model of purifying selection. 
The larger the $\beta$ more strongly the selection in nature favors these same, experimental preferences. 
Conversely, we interpret a $\beta<1$ as selection in nature favoring different amino acids than selection in lab and that the preferences are not a good description of natural evolution.

\ref{fig:compete} shows that as the divergence from the sequence used in the deep mutational scan increases, the value of $\beta$ decreases. 
This inverse relationship between $\beta$ and overall divergence is seen for the ExpCM's informed by both the H1 and H3 experiments.
Specifically, \jdbcomment{something about how the decreasing $\beta$ flattens the preferences and so loses information from the experiment.}
The flattening with increasing sequence divergence arises because the amino-acid preferences are increasingly less accurate as we move away from the experimental sequence on the tree.
As shown in \ref{fig:simulations}, the ability of models with site-specific stationary states to extend branch lengths requires these stationary states to be accurate\jdbcomment{Maybe expand the simulation figure to have two rows and have something with randomized preferences too}.
This fact also helps rationalize why the average of the H1 and H3 experiments does better across the whole tree \jdbcomment{refer to the figure showing that and give some explanation.}

The fact that amino-acid preferences shift as proteins evolve leaves us with an inherent tension: models with site-specific stationary states only become important for accurate branch-length estimation as sequences become increasingly diverged, but this same divergence degrades the accuracy of extrapolating the amino-acid preferences from any given experiment across the phylogenetic tree.
More importantly, the fact that amino-acid preferences shift over time means that there will not be any model with a single set of site-specific stationary states that accurately describes a phylogenetic tree that covers a wide span of sequences.
\jdbcomment{Use a slightly more succinct version of what you have below to drive home the point:
``This leaves us in a regime with inherent tension. 
Models which take into account the site-specific constraints of purifying selection in their stationary states are important for the accurate estimation of long branches but the site-specific constraints are not static throughout evolution. 
Models with a single stationary state regardless of lineage, site-specific or not, will not be able to capture the shifting functional constraints due to epistatic interactions.''}

>>>>

We can observe this phenomenon of shifting amino-acid preferences degrading the accuracy of the ExpCM by fitting trees containing increasingly diverged sequences.
Specifically, for each HA for which we have deep mutational scanning data, we created three HA phylogenetic trees: a ``low'' divergence tree that contains only sequences with \skhcomment{$\ge ?$} amino-acid identity to the HA used in the experiment, an ``intermediate'' divergence tree that contains sequences with \skhcomment{$\ge ?$} amino-acid identity to the HA in the experiment, and a ``high'' divergence tree that contains all the HAs (which have as little as 38\% identity to the HA in the experiment).
\ref{fig:compete} shows these sets of HA sequences.
For each of the subtrees in \ref{fig:compete}, we examined the congruence between site-specific natural selection over the tree and the amino-acid preferences measured in the deep mutational scanning experiment. 
We used the ExpCM fit parameter $\beta$ to measure this congruency~(\ref{eq:Frxy}). 
In essence, $\beta$ rescales the deep mutational scanning measurements~(\ref{fig:compete}). 
Larger values of $\beta$ indicate that natural selection prefers the same amino acids as the experiments but with a greater stringency, or a strong congruence between natural selection and the experimental preferences.

\ref{fig:compete} shows that as the divergence from the sequence used in the deep mutational scan increases, the value of $\beta$ decreases. 
This inverse relationship between $\beta$ and overall divergence is seen for the ExpCM's informed by both the H1 and H3 experiments.
Specifically, as the $\beta$ value decreases, the preferences ``flatten" and draw less information from the experiment. 
At the most extreme, $\beta = 0$, the preferences would be perfectly uniform and look identical to the GY94 preferences in \ref{fig:empirical_trees}. 
The flattening with increasing sequence divergence arises because the amino-acid preferences are increasingly less accurate as we move away from the experimental sequence on the tree.
As shown in \ref{fig:simulations}, the ability of models with site-specific stationary states to extend branch lengths requires these stationary states to be accurate\jdbcomment{Maybe expand the simulation figure to have two rows and have something with randomized preferences too}.
This fact also helps rationalize why the ExpCM defined by the average of the H1 and H3 experiments, representing the ``average" constraint on HA, does better across the whole tree (\ref{fig:empirical_trees}).

The fact that amino-acid preferences shift as proteins evolve leaves us with an inherent tension: models with site-specific stationary states only become important for accurate branch-length estimation as sequences become increasingly diverged, but this same divergence degrades the accuracy of extrapolating the amino-acid preferences from any given experiment across the phylogenetic tree.
More importantly, the fact that amino-acid preferences shift over time means that there will not be any model with a single set of site-specific stationary states that accurately describes a phylogenetic tree that covers a wide span of sequences.
